---
title: "OGC-biomass"
subtitle: "Environmental Comparisons"
author: "Kelly Murray Stoker, David Murray-Stoker"
output:
  pdf_document:
    toc: true
    toc_depth: 4
    fig_caption: yes
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(knitr.graphics.error = FALSE)
knitr::opts_chunk$set(results = "hold", fig.pos = "H", fig.align = "center", out.width = "92.5%")
kableExtra::usepackage_latex("float")
```

```{r Load Packages for R Markdown, include = FALSE}
library(kableExtra)
library(knitr)
```

```{r Load Packages, include = FALSE}
## Load the tidyverse
library(tidyverse)

## Load packages for data management and analysis
library(broom)
library(car)
library(easystats)
library(emmeans)
library(Kendall)
library(waterData)
```



\newpage
## Load Data

```{r Load Data, echo = TRUE}
## Read in sampling period environmental data
environmental.data <- read_csv(
  "data/OGC_final_biomass_data.csv",
  show_col_types = FALSE
) %>%
  select(
    UID, Year, Season, Period, CaCO, Carbon, Conductivity, DO, pH,
    Water_Temperature, Air_Temperature, Precipitation, Mean_Discharge
  )

## Set variables as factors
environmental.data$UID <- as_factor(environmental.data$UID)
environmental.data$Year <- as_factor(environmental.data$Year)
environmental.data$Season <- as_factor(environmental.data$Season)
environmental.data$Period <- as_factor(environmental.data$Period)

## Load the final workspace for reproducibility
load("data_analysis/OGC-biomass-environmental_comparisons-workspace.RData")
```



\newpage
# USGS Discharge Data

We obtained average daily discharge (m^3^ s^-1^) and water chemistry metrics from the United States Geological Survey (USGS, gage 02202500) to calculate the average discharge over a two-week period preceding each sampling date. Discharge data were downloaded using the `waterData` package. Water data were collected from the USGS water quality dataset, and air temperature and precipitation data were gathered from the National Weather Service Forecast Office in Louisville, GA.

\vspace{10pt}

```{r Water Data: Data Management, echo = TRUE}
## List of start dates
start.dates <- c(
  "1981-12-02", "1982-01-09", "1982-02-15", "1982-03-09",
  "1982-04-02", "1982-04-30", "1982-05-25", "1982-06-26",
  "1982-07-23", "1982-08-19", "1982-09-15", "1982-10-15",
  "1982-11-18", "1982-12-24", "1983-01-22", "1983-02-18",
  "1983-03-25", "1983-04-23", "1983-05-20", "1983-06-17",
  "1983-07-15", "1983-08-13", "1983-09-09", "1983-10-08",
  "1983-11-04",
  "2015-06-26", "2015-07-27", "2015-08-29", "2015-10-02",
  "2015-11-06", "2015-12-03", "2016-01-21", "2016-02-24",
  "2016-03-30", "2016-04-27", "2016-05-31", "2016-08-03",
  "2016-09-05", "2016-10-05", "2016-11-02", "2016-11-30",
  "2017-01-06", "2017-02-10", "2017-03-08", "2017-04-12",
  "2017-05-03", "2017-05-29", "2017-07-05", "2017-08-08"
)

## List of end dates
end.dates <- c(
  "1981-12-16", "1982-01-23", "1982-03-01", "1982-03-23",
  "1982-04-16", "1982-05-14", "1982-06-08", "1982-07-10",
  "1982-08-06", "1982-09-02", "1982-09-29", "1982-10-29",
  "1982-12-02", "1983-01-07", "1983-02-05", "1983-03-04",
  "1983-04-08", "1983-05-07", "1983-06-03", "1983-07-01",
  "1983-07-29", "1983-08-27", "1983-09-23", "1983-10-22",
  "1983-11-18",
  "2015-07-10", "2015-08-10", "2015-09-12", "2015-10-16",
  "2015-11-20", "2015-12-17", "2016-02-04", "2016-03-09",
  "2016-04-13", "2016-05-11", "2016-06-14", "2016-08-17",
  "2016-09-19", "2016-10-19", "2016-11-16", "2016-12-14",
  "2017-01-20", "2017-02-24", "2017-03-22", "2017-04-26",
  "2017-05-17", "2017-06-12", "2017-07-19", "2017-08-22"
)

## Bind start and end dates into a single dataframe
flow.dates <- as.data.frame(cbind(start.dates, end.dates))

## Add UID identifier to flow dates
flow.dates$UID <- environmental.data$UID
```

\newpage

```{r Water Data: Download, echo = TRUE, eval = FALSE}
## Empty dataframe for discharge data
discharge.data <- tibble(data.frame(matrix(0, nrow = 15, ncol = 49)))

## Append data for all dates
for (x in 1:49) {
  require(waterData)
  bin.1 <- importDVs(
    staid = "02202500", code = "00060", stat = "00003",
    sdate = flow.dates[x, 1],
    edate = flow.dates[x, 2]
  )
  bin.2 <- cleanUp(bin.1, task = "fix", replace = 0.001)
  bin.3 <- fillMiss(bin.1,
    block = 2, pmiss = 5, model = "trend",
    smooth = FALSE, log = "y"
  )
  bin.4 <- bin.3[, c(2, 3)]
  colnames(bin.4) <- c(flow.dates[x, 3])
  discharge.data[, x] <- bin.4[, 1]
}
```

\vspace{10pt}

```{r Water Data; Cleanup, echo = TRUE, eval = FALSE}
## Calculate two week mean & convert from ft3 to m3
mean.discharge <- sapply(discharge.data, mean)
metric.shift <- (mean.discharge / 35.315)

## Create mean discharge dataframe
metric.shift <- tibble(mean.discharge)
metric.shift$UID <- environmental.data$UID
```



\newpage
# Sampling Period Environmental Comparisons

\vspace{10pt}

## Environmental Variable ANOVAs

ANOVAs were conducted for discharge, precipitation, water temperature, and air temperature to test the effects of period, season, and the interaction with Type II sums-of-squares. ANOVA assumptions were inspected graphically using `check_model()`. Effect sizes for the ANOVAs were calculated as $\eta^2_P$ using `eta_squared()`. Pairwise contrasts between influential factors were performed using `emmeans()`, with effect sizes for contrasts calculated as Cohen's d.


\newpage
### Discharge

```{r Discharge ANOVA, echo = TRUE}
discharge.anova <- lm(
  log(Mean_Discharge) ~ Period * Season,
  data = environmental.data
)
```

\vspace{10pt}

```{r Discharge ANOVA: Check Assumptions, eval = FALSE}
check_model(discharge.anova)
```

\vspace{10pt}

```{r Discharge ANOVA: ANOVA Table, echo = FALSE}
kable(
  Anova(discharge.anova, type = "II"),
  booktabs = TRUE,
  digits = 3,
  caption = "ANOVA results for discharge by period, season, and the interaction.",
  col.names = c("Sums-of-Squares", "df", "F", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Discharge ANOVA: Effect Sizes, include = FALSE}
## Calculating partial eta-squared for each factor in the discharge ANOVA
discharge.anova.eta.squared <- eta_squared(
  Anova(discharge.anova, type = "II"),
  partial = TRUE
)
```

\vspace{10pt}

```{r Discharge ANOVA: Effect Size Table, echo = FALSE}
kable(
  discharge.anova.eta.squared,
  booktabs = TRUE,
  digits = 3,
  caption = "Table of the effect sizes in the discharge ANOVA.",
  col.names = c("Term", "Eta-squared", "CI", "CI Low", "CI High")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Calculate emmeans: Discharge ANOVA, include = FALSE}
## Calculate estimated marginal means
# Period
discharge.anova.emmeans.period <- emmeans(
  discharge.anova,
  specs = pairwise ~ Period,
  weights = "cells",
  adjust = "none"
)

# Season
discharge.anova.emmeans.season <- emmeans(
  discharge.anova,
  specs = pairwise ~ Season,
  weights = "cells",
  adjust = "none"
)

# Period x Season
discharge.anova.emmeans.period.season <- emmeans(
  discharge.anova,
  specs = pairwise ~ Period + Season,
  weights = "cells",
  adjust = "none"
)
```

\vspace{10pt}

```{r Estimated Marginal Means (Season): Discharge ANOVA, echo = FALSE}
kable(
  tidy(discharge.anova.emmeans.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by season in the discharge ANOVA.",
  col.names = c("Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Estimated Marginal Means (Period x Season): Discharge ANOVA, echo = FALSE}
kable(
  tidy(discharge.anova.emmeans.period.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by period and season in the discharge ANOVA.",
  col.names = c("Period", "Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Contrasts (Season): Discharge ANOVA, echo = FALSE}
kable(
  tidy(discharge.anova.emmeans.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by season in the discharge ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrasts (Period x Season): Discharge ANOVA, echo = FALSE}
kable(
  tidy(discharge.anova.emmeans.period.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by period and season in the discharge ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Season): Discharge ANOVA, echo = FALSE}
kable(
  eff_size(discharge.anova.emmeans.season,
    sigma = sigma(discharge.anova),
    edf = df.residual(discharge.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by season in the discharge ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Period x Season): Discharge ANOVA, echo = FALSE}
kable(
  eff_size(discharge.anova.emmeans.period.season,
    sigma = sigma(discharge.anova),
    edf = df.residual(discharge.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by period and season in the discharge ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```




\newpage
### Carbon

```{r Carbon ANOVA, echo = TRUE}
carbon.anova <- lm(
  Carbon ~ Period * Season,
  data = environmental.data
)
```

\vspace{10pt}

```{r Carbon ANOVA: Check Assumptions, eval = FALSE}
check_model(carbon.anova)
```

\vspace{10pt}

```{r Carbon ANOVA: ANOVA Table, echo = FALSE}
kable(
  Anova(carbon.anova, type = "II"),
  booktabs = TRUE,
  digits = 3,
  caption = "ANOVA results for carbon by period, season, and the interaction.",
  col.names = c("Sums-of-Squares", "df", "F", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Carbon ANOVA: Effect Sizes, include = FALSE}
## Calculating partial eta-squared for each factor in the carbon ANOVA
carbon.anova.eta.squared <- eta_squared(
  Anova(carbon.anova, type = "II"),
  partial = TRUE
)
```

\vspace{10pt}

```{r Carbon ANOVA: Effect Size Table, echo = FALSE}
kable(
  carbon.anova.eta.squared,
  booktabs = TRUE,
  digits = 3,
  caption = "Table of the effect sizes in the carbon ANOVA.",
  col.names = c("Term", "Eta-squared", "CI", "CI Low", "CI High")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Calculate emmeans: Carbon ANOVA, include = FALSE}
## Calculate estimated marginal means
# Period
carbon.anova.emmeans.period <- emmeans(
  carbon.anova,
  specs = pairwise ~ Period,
  weights = "cells",
  adjust = "none"
)

# Season
carbon.anova.emmeans.season <- emmeans(
  carbon.anova,
  specs = pairwise ~ Season,
  weights = "cells",
  adjust = "none"
)

# Period x Season
carbon.anova.emmeans.period.season <- emmeans(
  carbon.anova,
  specs = pairwise ~ Period + Season,
  weights = "cells",
  adjust = "none"
)
```

\vspace{10pt}

```{r Estimated Marginal Means (Season): Carbon ANOVA, echo = FALSE}
kable(
  tidy(carbon.anova.emmeans.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by season in the carbon ANOVA.",
  col.names = c("Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Estimated Marginal Means (Period x Season): Carbon ANOVA, echo = FALSE}
kable(
  tidy(carbon.anova.emmeans.period.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by period and season in the carbon ANOVA.",
  col.names = c("Period", "Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Contrasts (Season): Carbon ANOVA, echo = FALSE}
kable(
  tidy(carbon.anova.emmeans.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by season in the carbon ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrasts (Period x Season): Carbon ANOVA, echo = FALSE}
kable(
  tidy(carbon.anova.emmeans.period.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by period and season in the carbon ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Season): Carbon ANOVA, echo = FALSE}
kable(
  eff_size(carbon.anova.emmeans.season,
    sigma = sigma(carbon.anova),
    edf = df.residual(carbon.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by season in the carbon ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Period x Season): Carbon ANOVA, echo = FALSE}
kable(
  eff_size(carbon.anova.emmeans.period.season,
    sigma = sigma(carbon.anova),
    edf = df.residual(carbon.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by period and season in the carbon ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```


\newpage
### Precipitation

```{r Precipitation ANOVA, echo = TRUE}
precipitation.anova <- lm(
  Precipitation ~ Period * Season,
  data = environmental.data
)
```

\vspace{10pt}

```{r Precipitation ANOVA: Check Assumptions, eval = FALSE}
check_model(precipitation.anova)
```

\vspace{10pt}

```{r Precipitation ANOVA: ANOVA Table, echo = FALSE}
kable(
  Anova(precipitation.anova, type = "II"),
  booktabs = TRUE,
  digits = 3,
  caption = "ANOVA results for precipitation by period, season, and the interaction.",
  col.names = c("Sums-of-Squares", "df", "F", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Precipitation ANOVA: Effect Sizes, include = FALSE}
## Calculating partial eta-squared for each factor in the precipitation ANOVA
precipitation.anova.eta.squared <- eta_squared(
  Anova(precipitation.anova, type = "II"),
  partial = TRUE
)
```

\vspace{10pt}

```{r Precipitation ANOVA: Effect Size Table, echo = FALSE}
kable(
  precipitation.anova.eta.squared,
  booktabs = TRUE,
  digits = 3,
  caption = "Table of the effect sizes in the precipitation ANOVA.",
  col.names = c("Term", "Eta-squared", "CI", "CI Low", "CI High")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Calculate emmeans: Precipitation ANOVA, include = FALSE}
## Calculate estimated marginal means
# Period
precipitation.anova.emmeans.period <- emmeans(
  precipitation.anova,
  specs = pairwise ~ Period,
  weights = "cells",
  adjust = "none"
)

# Season
precipitation.anova.emmeans.season <- emmeans(
  precipitation.anova,
  specs = pairwise ~ Season,
  weights = "cells",
  adjust = "none"
)

# Period x Season
precipitation.anova.emmeans.period.season <- emmeans(
  precipitation.anova,
  specs = pairwise ~ Period + Season,
  weights = "cells",
  adjust = "none"
)
```

\vspace{10pt}

```{r Estimated Marginal Means (Season): Precipitation ANOVA, echo = FALSE}
kable(
  tidy(precipitation.anova.emmeans.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by season in the precipitation ANOVA.",
  col.names = c("Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Contrasts (Season): Precipitation ANOVA, echo = FALSE}
kable(
  tidy(precipitation.anova.emmeans.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by season in the precipitation ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Season): Precipitation ANOVA, echo = FALSE}
kable(
  eff_size(precipitation.anova.emmeans.season,
    sigma = sigma(precipitation.anova),
    edf = df.residual(precipitation.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by season in the precipitation ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```


\newpage
### Water Temperature

```{r Water Temperature ANOVA, echo = TRUE}
water.temperature.anova <- lm(
  Water_Temperature ~ Period * Season,
  data = environmental.data
)
```

\vspace{10pt}

```{r Water Temperature ANOVA: Check Assumptions, eval = FALSE}
check_model(water.temperature.anova)
```

\vspace{10pt}

```{r Water Temperature ANOVA: ANOVA Table, echo = FALSE}
kable(
  Anova(water.temperature.anova, type = "II"),
  booktabs = TRUE,
  digits = 3,
  caption = "ANOVA results for water temperature by period, season, and the interaction.",
  col.names = c("Sums-of-Squares", "df", "F", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Water Temperature ANOVA: Effect Sizes, include = FALSE}
## Calculating partial eta-squared for each factor in the water temperature ANOVA
water.temperature.anova.eta.squared <- eta_squared(
  Anova(water.temperature.anova, type = "II"),
  partial = TRUE
)
```

\vspace{10pt}

```{r Water Temperature ANOVA: Effect Size Table, echo = FALSE}
kable(
  water.temperature.anova.eta.squared,
  booktabs = TRUE,
  digits = 3,
  caption = "Table of the effect sizes in the water temperature ANOVA.",
  col.names = c("Term", "Eta-squared", "CI", "CI Low", "CI High")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Calculate emmeans: Water Temperature ANOVA, include = FALSE}
## Calculate estimated marginal means
# Period
water.temperature.anova.emmeans.period <- emmeans(
  water.temperature.anova,
  specs = pairwise ~ Period,
  weights = "cells",
  adjust = "none"
)

# Season
water.temperature.anova.emmeans.season <- emmeans(
  water.temperature.anova,
  specs = pairwise ~ Season,
  weights = "cells",
  adjust = "none"
)

# Period x Season
water.temperature.anova.emmeans.period.season <- emmeans(
  water.temperature.anova,
  specs = pairwise ~ Period + Season,
  weights = "cells",
  adjust = "none"
)
```

\vspace{10pt}

```{r Estimated Marginal Means (Season): Water Temperature ANOVA, echo = FALSE}
kable(
  tidy(water.temperature.anova.emmeans.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by season in the water temperature ANOVA.",
  col.names = c("Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Contrasts (Season): Water Temperature ANOVA, echo = FALSE}
kable(
  tidy(water.temperature.anova.emmeans.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by season in the water temperature ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Season): Water Temperature ANOVA, echo = FALSE}
kable(
  eff_size(water.temperature.anova.emmeans.season,
    sigma = sigma(water.temperature.anova),
    edf = df.residual(water.temperature.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by season in the water temperature ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```


\newpage
### Air Temperature

```{r Air Temperature ANOVA, echo = TRUE}
air.temperature.anova <- lm(
  Air_Temperature ~ Period * Season,
  data = environmental.data
)
```

\vspace{10pt}

```{r Air Temperature ANOVA: Check Assumptions, eval = FALSE}
check_model(air.temperature.anova)
```

\vspace{10pt}

```{r Air Temperature ANOVA: ANOVA Table, echo = FALSE}
kable(
  Anova(air.temperature.anova, type = "II"),
  booktabs = TRUE,
  digits = 3,
  caption = "ANOVA results for air temperature by period, season, and the interaction.",
  col.names = c("Sums-of-Squares", "df", "F", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Air Temperature ANOVA: Effect Sizes, include = FALSE}
## Calculating partial eta-squared for each factor in the air temperature ANOVA
air.temperature.anova.eta.squared <- eta_squared(
  Anova(air.temperature.anova, type = "II"),
  partial = TRUE
)
```

\vspace{10pt}

```{r Air Temperature ANOVA: Effect Size Table, echo = FALSE}
kable(
  air.temperature.anova.eta.squared,
  booktabs = TRUE,
  digits = 3,
  caption = "Table of the effect sizes in the air temperature ANOVA.",
  col.names = c("Term", "Eta-squared", "CI", "CI Low", "CI High")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Calculate emmeans: Air Temperature ANOVA, include = FALSE}
## Calculate estimated marginal means
# Period
air.temperature.anova.emmeans.period <- emmeans(
  air.temperature.anova,
  specs = pairwise ~ Period,
  weights = "cells",
  adjust = "none"
)

# Season
air.temperature.anova.emmeans.season <- emmeans(
  air.temperature.anova,
  specs = pairwise ~ Season,
  weights = "cells",
  adjust = "none"
)

# Period x Season
air.temperature.anova.emmeans.period.season <- emmeans(
  air.temperature.anova,
  specs = pairwise ~ Period + Season,
  weights = "cells",
  adjust = "none"
)
```

\vspace{10pt}

```{r Estimated Marginal Means (Season): Air Temperature ANOVA, echo = FALSE}
kable(
  tidy(air.temperature.anova.emmeans.season$emmeans),
  booktabs = TRUE,
  digits = 3,
  caption = "Estimated marginal means by season in the air temperature ANOVA.",
  col.names = c("Season", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\vspace{10pt}

```{r Contrasts (Season): Air Temperature ANOVA, echo = FALSE}
kable(
  tidy(air.temperature.anova.emmeans.season$contrasts),
  booktabs = TRUE,
  digits = 3,
  caption = "Pairwise contrasts by season in the air temperature ANOVA.",
  col.names = c("Term", "Contrast", "Null Value", "Estimate", "SE", "df", "t", "P-value")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```

\vspace{10pt}

```{r Contrast Effect Sizes (Season): Air Temperature ANOVA, echo = FALSE}
kable(
  eff_size(air.temperature.anova.emmeans.season,
    sigma = sigma(air.temperature.anova),
    edf = df.residual(air.temperature.anova)
  ),
  booktabs = TRUE,
  digits = 3,
  caption = "Effect sizes for the pairwise contrasts by season in the air temperature ANOVA.",
  col.names = c("Contrast", "Cohen's d", "SE", "df", "CI Lower", "CI Upper")
) %>%
  kable_styling(latex_options = c("HOLD_position", "striped", "scale_down"))
```


\newpage
# Long-Term Environmental Comparisons

## Data Management

We obtained historical USGS discharge data from December 1969 to November 2018 to determine average daily discharge for the winter-spring season of each year. Water data collected from 1974 until 2018 were also obtained from the USGS water quality dataset. Precipitation and air temperature data from 1970 until 2018 were obtained from the National Weather Service Forecast Office in Louisville, GA. We then filtered the long-term environmental data to only include the winter-spring season group (winter months = December, January, and February; spring months = March, April, and May), which are important for the flood pulse into the Ogeechee River. Mean values were then aggregated by year to generate a mean winter-spring value per year for use in time series analyses. Trends in the the time series were analyzed with Mann-Kendall tests using `MannKendall()`.

\vspace{10pt}

```{r Load Long-Term Environmental Data, echo = TRUE}
## Read in long-term environmental data
long.term.environmental.data <- read_csv(
  "data/OGC_long_term_environmental_data.csv",
  show_col_types = FALSE
)

## Subset data by season groups
winter.spring.data <- long.term.environmental.data %>%
  filter(Season_Group == "WinterSpring")
```

\vspace{10pt}

```{r Data Management: Winter-Spring, echo = TRUE}
## Aggregate mean values by year for the winter-spring data

## Winter-spring discharge
winter.spring.discharge <- aggregate(
  Mean_Discharge ~ Year,
  data = winter.spring.data,
  FUN = mean
)

## Winter-spring precipitation
winter.spring.precipitation <- aggregate(
  Precipitation ~ Year,
  data = winter.spring.data,
  FUN = mean
) %>%
  na.omit()

## Winter-spring water temperature
winter.spring.water.temperature <- aggregate(
  Water_Temperature ~ Year,
  data = winter.spring.data,
  FUN = mean
) %>%
  na.omit()

## Winter-spring air temperature
winter.spring.air.temperature <- aggregate(
  Air_Temperature ~ Year,
  data = winter.spring.data,
  FUN = mean
) %>%
  na.omit()
```



\newpage
## Mann-Kendall Tests on Winter-Spring Trends

We assessed temporal trends in discharge, precipitation, water temperature, and air temperature using Mann-Kendall tests for the winter-spring season group (i.e., flood-prone seasons). We calculated monthly averages for each variable, and then assessed a shift in the time series using `MannKendall()`.

\vspace{10pt}

```{r Winter-Spring: Discharge, results = FALSE}
MannKendall(winter.spring.discharge$Mean_Discharge)
# tau = -0.248, P = 0.012
```

\vspace{10pt}

```{r Winter-Spring: Precipitation, results = FALSE}
MannKendall(winter.spring.precipitation$Precipitation)
# tau = -0.087, P = 0.384
```

\vspace{10pt}

```{r Winter-Spring: Air Temperature, results = FALSE}
MannKendall(winter.spring.air.temperature$Air_Temperature)
# tau = -0.11, P = 0.270
```

\vspace{10pt}

```{r Winter-Spring: Water Temperature, results = FALSE}
MannKendall(winter.spring.water.temperature$Water_Temperature)
# tau = 0.442, P < 0.001
```





```{r Save Workspace, include = FALSE}
## Save workspace
save.image("data_analysis/OGC-biomass-environmental_comparisons-workspace.RData")
```



\newpage
# R Session Information


```{r R Packages, echo = FALSE}
df_session_packages <- devtools::session_info()$packages %>%
  as.data.frame(.) %>%
  filter(attached == TRUE) %>%
  dplyr::select(loadedversion, date) %>%
  rownames_to_column()

colnames(df_session_packages) <- c("Package", "Loaded Version", "Date")

kable(
  df_session_packages,
  booktabs = TRUE,
  caption = "Packages for data management and analyses."
) %>%
  kable_styling(
    full_width = FALSE,
    latex_options = c("HOLD_position", "striped")
  )
```
